Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2019-07-04 unimodular-covers.tex
put back two questions I accidentally removed
worked on the proof in the appendix

\renewcommand{\int}{\operatorname{int}}
\newcommand{\width}{\operatorname{width}}
\newcommand{\Vol}{\operatorname{Vol}}
\newcommand{\length}{\operatorname{length}}

@@

At some point I'll assume volume two for the parallelogram $p_1p_2 + q_1q_2$, but for the time being we don't.

Suppose by contradiction that there is no lattice point as described in the lemma. Then there cannot be a vertex of $P$ in the corresponding strip, since $P$ is a lattice polygon. The same holds for $Q$ and its corresponding strip. Thus $P$ has two primitive edges which each have one vertex on each side of the strip $(p_1p_2 + \langle \overrightarrow{q_1q_2}\rangle)$; we will call these edges $\ell$ and $r$ (with $\ell$ to the left and $r$ to the right). The same holds for $Q$ and the strip $(q_1q_2 + \langle \overrightarrow{p_1p_2}\rangle)$, and we call the edges  $b, t$ (with $b$ below and $t$ above). To make things more explicit, we assume the following coordinates for the edge-vectors:
\[
\vec l = (l_1,l_2), \quad
\vec r= (r_1,r_2), \quad
\vec b= (b_1,b_2), \quad
\vec t= (t_1,t_2).
\]
Without loss of generality we assume $l_2,r_2,b_1,t_1 >0$. (They are non-zero and we can arbitrarily choose a sign for each vector).

@@

\begin{claim}
$l_2, r_2, b_1, t_1 >1$.
\end{claim}

\begin{proof}
$l_2, r_2, b_1, t_1 \ge 1$ follows from the fact that each edge crosses the corresponding strip. 
If one of the inequalities is not strict, then one edge must have vertices on the corresponding strip. Then one of $(p_1p_2 + \langle \overrightarrow{q_1q_2}\rangle) \cap P$ or $(q_1q_2 + \langle \overrightarrow{p_1p_2}\rangle) \cap Q$
contains a triangle with one edge a translation of $p_1p_2$ and another edge a translation of $q_1q_2$. Such a triangle must contain a lattice point other than its vertices, and since $p_1p_2$ and $q_1q_2$ are primitive, the lattice point must be in the interior of the strip.
\end{proof}

\begin{claim}
$b_2$ and $t_2$ are non-zero and have the same sign. W.l.o.g.~we assume they are positive.
\end{claim}

\begin{proof}
If, say, $b_2 \le 0 \le t_2$. Then the intersections of $t$ and $b$ with the vertical line $\{x=-\beta\}$ through $q_1$ are less than distance one apart (because $t$ must go below every lattice point in the interior of $p_1p_2 + q_1q_2 - p_1$ and $b$ above every latttice point in the interior of $p_1p_2 + q_1q_2 - p_2$). This implies the width w.r.t to $f_q$, the functional constant on $q_1q_2$, of any segment with vertices in $Q \cap \{x\le  -\beta\}$ is less than one, so it is impossible to fit a translated copy of $l$ in it.
\end{proof}

\begin{claim}
Assume w.l.o.g.~that $b$ has greater slope than $t$. Then, 
\begin{enumerate}
\item The intersection of $Q$ with any vertical line to the right of $q_2$ has length strictly smaller than one.
\item $r_1>0$.
\end{enumerate}
\end{claim}

\begin{proof}
The assumption implies that the intersections of $t$ and $b$ with the vertical line $\{x=1-\beta\}$ through $q_2$ are less than distance one apart (consider a vertical lattice line intersecting $q_1q_2$ and two consecutive lattice points in it, one above and one below the horizontal axis). This in turn implies part (a).

For (b), if $r_1$ was $\le 0$ it would be impossible to fit a translated copy of $r$ in the right side of $Q$, because of (a).
\end{proof}

@@

Starting here we assume index 2.

\begin{claim}
The slopes of $t$ and $b$ are strictly smaller than one and that of $r$ strictly greater than one.
\end{claim}

\begin{proof}
Otherwise the points $(0,1/2)$, $(0,-1/2)$, $(1/2,0)$, respectively, would be in the corresponding band.
\end{proof}

We are now ready to show a contradiction. Let $u$ and $v$ be the end-points of the translated copy of $r$ in $Q$, with $\vec r=v-u$. Let $d$ be the line of slope $1$ through $u$, which is a lattice line. Let $d'$ be the next parallel lattice line above $d$. By Claim A.3.1, the intersection of $d'$ with the vertical line through $u$ is above $Q$. Since the slope of $t$ is less than one, the same happens for every point of $d'$ to the right of $u$. In particular, $v$ must be below $d'$ and hence it must be on or below $d$. This contradicts the fact that the slope of $r$ is strictly greater than one. QED

>>>>

\renewcommand{\int}{\operatorname{int}}
\newcommand{\width}{\operatorname{width}}
\newcommand{\ind}{\operatorname{index}}
\newcommand{\Vol}{\operatorname{Vol}}
\newcommand{\length}{\operatorname{length}}

@@

%At some point I'll assume volume two for the parallelogram $p_1p_2 + q_1q_2$, but for the time being we don't.

Let us fix coordinates in which $p=\conv(p_1,p_2)$ and $q= \conv(q_1,q_2)$ are a vertical and horizontal segment of length one intersecting at the origin. That is, there are numbers $\alpha, \beta \in (0,1)$ such that
\[
p_1=(0,-\alpha), \quad
p_2=(0,1-\alpha), \quad
q_1=(-\beta,0), \quad
q_2=(1-\beta,0).
\] \giulia{perhaps the explicit coordinates are superfluous now}
The ambient lattice is a certain lattice containing $p_1,p_2,q_1,q_2$ (the case of index two is when $\alpha=\beta=1/2$ and the lattice is spanned by these four points). We can further assume (since translating the polytopes by lattice vectors will not result in any loss of generality) that $p_1$ is on the consecutive lattice line parallel to the line $q$ lies on. That is, if $f_q$ is the primitive lattice functional which is zero on $q$, then $f_q(p_1)=-1$. Equivalently, the triangle $\conv(q_1, q_2, p_1)$ is unimodular. Let $f_p$ be the functional constant on $p$. We observe then that 
\[
\width_{f_q}(p + \overrightarrow{q})=\width_{f_q}(p)=\width_{f_p}(q)=\width_{f_p}(q +\overrightarrow{p}) = \ind(p, q) =:i.
\]

Observe that $p_1$ is the unique lattice point in the strip $q +\overrightarrow{p}$ on the line $f_q(x)=-1$. Indeed, since $q$ is primitive, the only way that in the strip there could be two lattice points on $f_q(x)=-1$ is if they were on the boundary of the strip, which would however imply that $p$ and $q$ have index $1$, which is a contradiction. The unique lattice point in the strip on the line $f_q(x)=1$ is then $q_1+q_2 -p_1$.

Suppose by contradiction that there is no lattice point as described in the lemma. Then there cannot be a vertex of $P$ in the corresponding strip, since $P$ is a lattice polygon. The same holds for $Q$ and its corresponding strip. Thus $P$ has two primitive edges which each have one vertex on each side of the strip $(p_1p_2 + \langle \overrightarrow{q_1q_2}\rangle)$; we will call these edges $\ell$ and $r$ (with $\ell$ to the left and $r$ to the right). The same holds for $Q$ and the strip $(q_1q_2 + \langle \overrightarrow{p_1p_2}\rangle)$, and we call the edges  $b, t$ (with $b$ below and $t$ above).  Observe that one of $l$ and $r$ can coincide with $p$, if this is an edge of $P$, and similarly one of $t,b$ might be $q$, if this is an edge of $Q$. To make things more explicit, we assume the following coordinates for the edge-vectors:
\[
\vec l = (l_1,l_2), \quad
\vec r= (r_1,r_2), \quad
\vec b= (b_1,b_2), \quad
\vec t= (t_1,t_2).
\]
Without loss of generality we assume $l_2,r_2,b_1,t_1 >0$. (They are non-zero and we can arbitrarily choose a sign for each vector).

@@

%\begin{claim}
%$l_2, r_2, b_1, t_1 >1$.
%\end{claim}

\begin{claim}\giulia{I think we need to treat separately the case where we deal with edges (these are easier, in any case, but I couldn't find a unification)}
The following inequalities hold, provided the segment in question is not an edge of the corresponding polytope:
\begin{align*}
\width_{f_q}(l) > i, \\
\width_{f_q}(r) > i, \\
\width_{f_p}(t) > i, \\
\width_{f_p}(b) > i.
\end{align*}
\end{claim}

\begin{proof}
The inequality $\geq i$ follows in each case from the fact that each segment crosses the corresponding strip, the width of which (w.r.t the appropriate functional) is $i$, as observed above. 
If one of the inequalities were not strict, then the corresponding segment, say $l$, must have vertices on the corresponding strip. Then the strip $(p + \langle \overrightarrow{q}\rangle) \cap P$ contains a triangle with one edge a translation of $p$ and another edge a translation of $q$. Such a triangle must contain a lattice point other than its vertices, and since $p_1p_2$ and $q_1q_2$ are primitive, the lattice point must be in the interior of the strip.
\end{proof}

%\begin{claim}
%$b_2$ and $t_2$ are non-zero and have the same sign. W.l.o.g.~we assume they are positive.
%\end{claim}
%
%\begin{proof}
%If, say, $b_2 \le 0 \le t_2$. Then the intersections of $t$ and $b$ with the vertical line $\{x=-\beta\}$ through $q_1$ are less than distance one apart (because $t$ must go below every lattice point in the interior of $p_1p_2 + q_1q_2 - p_1$ and $b$ above every latttice point in the interior of $p_1p_2 + q_1q_2 - p_2$). This implies the width w.r.t to $f_q$, the functional constant on $q_1q_2$, of any segment with vertices in $Q \cap \{x\le  -\beta\}$ is less than one, so it is impossible to fit a translated copy of $l$ in it.
%\end{proof}

\begin{claim}
The slopes of $b$ and $t$ must either be both strictly larger than the slope of $q$ or both be strictly smaller. W.l.o.g.~we assume they are larger.
\end{claim}

\begin{proof}
If, say, $b$ has positive slope and $t$ negative. Then since $t$ intersects the vertical line $x=0$ spanned by $p$ below $p_2$, and $b$ intersects it above $p_1$, the slopes of $b$ and $t$ imply that  $Q \cap \{x\ge 0\}$ is contained in the strip $-1<f_q(x)<i-1$, of width $i$. This cannot contain a translated copy of $r$, since $\width_{f_q}(r) >i$ .
\end{proof}

\begin{claim}
\label{positive slopes}
Assume w.l.o.g.~that $b$ has greater slope than $t$. Then, 
\begin{enumerate}
\item The intersection of $Q$ with any vertical line to the right of $p$ has width w.r.t. $f_q$ strictly smaller than $i$.
\item $r$ also has positive slope.
\end{enumerate}
\end{claim}

\begin{proof}
\giulia{translate this into the width language!}
Since $t$ and $b$ both intersect $p$, these intersections form a vertical segment which has width w.r.t $f_q$ less than $i$, the width of $p$. The assumptions on the slopes implies that this is true for any vertical segment with endpoints in $b$ and $t$ to the right of this one, which is part (a).

For (b), if $r$ had negative slope, it would be impossible to fit a translated copy of $r$ in the right side of $Q$, because of (a): it would lie \giulia{finish, with picture}
\end{proof}

@@

%Starting here we assume index 2.
%
%\begin{claim}
%The slopes of $t$ and $b$ are strictly smaller than one and that of $r$ strictly greater than one.
%\end{claim}
%
%\begin{proof}
%Otherwise the points $(0,1/2)$, $(0,-1/2)$, $(1/2,0)$, respectively, would be in the corresponding band.
%\end{proof}
%
%We are now ready to show a contradiction. Let $u$ and $v$ be the end-points of the translated copy of $r$ in $Q$, with $\vec r=v-u$. Let $d$ be the line of slope $1$ through $u$, which is a lattice line. Let $d'$ be the next parallel lattice line above $d$. By Claim A.3.1, the intersection of $d'$ with the vertical line through $u$ is above $Q$. Since the slope of $t$ is less than one, the same happens for every point of $d'$ to the right of $u$. In particular, $v$ must be below $d'$ and hence it must be on or below $d$. This contradicts the fact that the slope of $r$ is strictly greater than one. QED

\begin{claim}
The slopes of $t$ and $b$ are strictly smaller than the slope of $\conv(p_1, q_2)$ and that of $r$ (and $l$) is strictly greater than it.
\end{claim}

\begin{proof}
Since $b$ and $t$ must respectively separate $p_1$ and $q_1+q_2-p_1$ from the remaining vertices of the parallelogram $\conv(q_1, p_1, q_2, q_1+q_2-p_1)$, they must respectively intersect its (parallel) edges $\conv(p_1, q_2)$ and $\conv(q_1, q_1+q_2-p_1)$. This means their slopes must be strictly less than that of these segments. The same argument can be applied to $l$ and $r$ if one considers the parallelogram  $\conv(p_1, q_2, p_2, p_1+p_2-p_2)$, yielding that the slopes of $l$ and $r$ are strictly larger than that of $\conv(p_1, q_2)$.
\end{proof}

We are now ready to show a contradiction. Since $Q$ has normal fan refining that of $P$, it must have an edge which is a translated copy of $r$. Let $u$ and $v$ be its endpoints. Now consider the lattice line $d$ through $u$ parallel to  $\conv(p_1, q_2)$. Let $d'$ be the next parallel lattice line above $d$. The width w.r.t $f_q$ of the vertical segment starting at $u$ with other endpoint $v'$ on $d'$ is $i$, since this vertical segment is a translation of $p$. Thus $v'$ must lie outside $Q$, by claim \ref{positive slopes}. That is, $v'$ lies above the line spanned by $t$. Since the slope of $t$ is less than the slope of $d$, the same happens for every point of $d'$ to the right of $u$. In particular, $v$ must be below $d'$ and hence it must be on or below $d$. This contradicts the fact that the slope of $r$ is strictly greater than the slope of $d$.
